---
output: github_document
always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
library(vtable)
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# vtable

<!-- badges: start -->

[![](http://cranlogs.r-pkg.org/badges/grand-total/vtable?color=orange)](https://cran.r-project.org/package=vtable) [![](http://cranlogs.r-pkg.org/badges/last-month/vtable?color=green)](https://cran.r-project.org/package=vtable) [![](https://www.r-pkg.org/badges/version/vtable?color=blue)](https://cran.r-project.org/package=vtable) [![CRAN
checks](https://cranchecks.info/badges/summary/vtable)](https://cran.r-project.org/web/checks/check_results_vtable.html) [![CRAN status](https://www.r-pkg.org/badges/version/vtable)](https://CRAN.R-project.org/package=vtable)

<!-- badges: end -->

The **vtable** package is designed to help you quickly and efficiently look at and document your data.

There are three main functions in **vtable**:

1. `vtable`, or `vt` for short, shows you information about the variables in your data set, including variable labels, in a way that is easy to use "find in page" to search through. It was designed to be similar to Stata's "Variables" panel.
2. `sumtable` or `st` for short, provides a table of summary statistics. It is very similar in spirit to the summary statistics function of `stargazer::stargazer()` except that it accepts `tibble`s, handles factor variables, and makes by-group statistics and group tests easy.
3. `labeltable` provides a table of value labels, either for variables labelled with **sjlabelled** or **haven** or similar, or for when you want to see how the values of one column line up with the values of another.

All three of these functions are built with the intent of being *fast*. Not so much fast to *run*, but fast to *use*. The defaults are intended to be good defaults, and the output by default prints to the Viewer tab (in RStudio) or the browser (outside RStudio) so you can see it immediately, and continue to look at it as you work on your data. 

You could almost certainly build your own highly-customized version of `vtable`, But why do that when you can just do `vt(df)` and see the information you need to see? And there are eight million packages that make summary statistics tables to your exact specifications if you tweak them. But there's a good chance that `st(df)` does what you want. If you want something real out there, that's when you can break out the big guns.

All three main **vtable** functions can produce HTML, LaTeX, `data.frame`, CSV, or `knitr::kable()` output.

## Installation

You can install vtable from CRAN. Note that the documentation on this site refers to the development version, and so may not work perfectly for the CRAN version. But the two will usually be the same.: 

```r
install.packages("vtable")
```

### Development version 

The development version can be installed from [GitHub](https://github.com/):

``` r
# install.packages("remotes")
remotes::install_github("NickCH-K/vtable")
```
## vtable Example

I'll just do a brief example here, using the `iris` we all know and love. Output will be to `kable` since this is an RMarkdown document.

```{r}
data(iris)

# Basic vtable
vt(iris)
```

There are plenty of options if we want to go nuts, but let's keep it simple and just ask for a little more with `lush`

```{r}
vt(iris, lush = TRUE)
```

## sumtable Example

Let's stick with `iris`!

```{r}
# Basic summary stats
st(iris)
```

Note that `sumtable` allows for much more customization than `vtable` since there's a heightened chance you want it for a paper or something. But I'll leave that to the more detailed documentation. For now just note it does by-group stats, either in "`group.long`" format (multiple `sumtable`s stacked on top of each other), or by default, in columns, with an option to add a group test.

Grouped `sumtables` look a little nicer in formats that suport multi-column cells like HTML and LaTeX. 

**These tables include multi-column cells, which are not supported in the `kable` output, but are supported by `vtable`'s `dftoHTML` and `dftoLaTeX` functions. They look nicer in the HTML or LaTeX output.**

```{r}
st(iris, 
   group = 'Species', 
   group.test = TRUE)
```

## labeltable Example

For this we'll need labeled values. 

```{r}
data(efc, package = 'sjlabelled')

# Now shoot - how was gender coded?
labeltable(efc$e16sex)
```

`labeltable` can also be used to see, for values of one variable, what values are present of other variables. This is intended for use if one variable is a recode, simplification, or lost-labels version of another, but hey, go nuts.

```{r}
labeltable(efc$e15relat,efc$e16sex,efc$e42dep)
```
