---
title: "Variable Table"
author: "Nick Huntington-Klein"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
<!-- output: rmarkdown::html_vignette. pdf_document -->
vignette: >
  %\VignetteIndexEntry{Variable Table (vtable)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The `vtable` package serves the purpose of outputting automatic variable documentation that can be easily viewed while continuing to work with data.

`vtable` contains three functions: `vtable()`, `labeltable()`, and `dftoHTML()`. `vtable()` takes a dataset and outputs a formatted variable documentation file. This serves several purposes. 

First, it allows for an easy generation of a variable documentation file, without requiring that one has already been created and made accessible through `help(data)`, or dealing with creating and finding R help documentation files. 

Second, it produces a list of variables (and, if provided, their labels) that can be easily viewed while working with the data, preventing repeated calls to `head()`, and making it much easier to work with confusingly-named variables. 

Third, the variable documentation file can be opened in a browser (with option `out='browser'`, saving to file and opening directly, or by opening in the RStudio Viewer pane and clicking 'Show in New Window') where it can be easily searched with standard Find-in-Page functions like Ctrl/Cmd-F, allowsing you to search for the variable or variable label you want.

`labeltable()` is designed to take a single variable and show the values it is associated with. This can also be used to generate data documentation if desired, or can just be an easy way to look at label values, or learn more about the data you're working with.

If that variable has value labels from the `sjlabelled` or `haven` packages, it will show how the values in the data correspond to the value labels.

Alternately, you can include other variables as well, and `labeltable()` will show, for each value of the variable you're interested in, the values that those other variables take. This can be handy, for example, if you used some variables to create a numeric ID and want to remember what original values correspond to each ID. It can also act as sort of a cross-tabulation.

`dftoHTML()` is a helper function used by `vtable()` and `labeltable()`. It takes any data frame or matrix with column names and outputs HTML table code for that data.

-----

# The `vtable()` function

`vtable()` syntax follows the following outline:

```{r, eval=FALSE}
vtable(data, 
    out=NA,
    file=NA, 
    labels=NA, 
    class=TRUE, 
    values=TRUE,
    missing=FALSE,
    index=FALSE,
    factor.limit=5,
    char.values=FALSE,
    data.title=NA,
    desc=NA, 
    col.width=NA,
    summ=NA)
```

The goal of `vtable()` is to take a data set `data` and output an HTML file with documentation concerning each of the variables in `data`. There are several options as to what will be included in the documentation file, and each of these options are explained below. Throughout, most code will be included without output because output is typically rendered in a browser or in the Viewer pane of RStudio, not the console:

## `data`

The `data` argument can take any data frame, data table, tibble, or matrix, as long as it has a valid set of variable names stored in the `colnames()` attribute. The goals of `vtable()` is to produce documentation of each of the variables in this data set and display that documentation, one variable per row on the output `vtable`.

If `data` has embedded variable or value labels, as the data set `efc` does below, `vtable()` will extract and use them automatically.
                   
```{r, eval=FALSE}
library(vtable)

#Example 1, using base data LifeCycleSavings
data(LifeCycleSavings)
vtable(LifeCycleSavings)

#Example 2, using constructed data frame
df <- data.frame(var1 = 1:4, var2 = c('A','B','C','D'))
vtable(df)

#Example 3, using matrix with column names
matrix <- as.matrix(df)
vtable(df)

#Example 4, using efc data with embedded variable labels
library(sjlabelled)
data(efc)
vtable(efc)
```

## `out`

The `out` option determines what will be done with the resulting variable documentation file. There are several options for `out`:

| Option     | Result |
|------------| -----------------------------------------|
| browser    | Loads variable documentation in web browser. |
| viewer     | Loads variable documentation in Viewer pane (RStudio only). |
| htmlreturn | Returns HTML code for variable documentation file. |
| return     | Returns variable documentation table in data frame format. |

By default, `vtable` will select 'viewer' if running in RStudio, and 'browser' otherwise.

```{r, eval=FALSE}
library(vtable)

data(LifeCycleSavings)
vtable(LifeCycleSavings)
vtable(LifeCycleSavings,out='browser')
vtable(LifeCycleSavings,out='viewer')
htmlcode <- vtable(LifeCycleSavings,out='htmlreturn')
vartable <- vtable(LifeCycleSavings,out='return')
```


## `file`

The `file` argument will write the variable documentation file to an HTML file and save it. Will automatically append 'html' filetype if the filename does not include a period.

```{r, eval=FALSE}
library(vtable)

data(LifeCycleSavings)
vtable(LifeCycleSavings,file='lifecycle_variabledocumentation')
```

## `labels`

The `labels` argument will attach variable labels to the variables in `data`. If variable labels are embedded in `data` and those labels are what you want, the `labels` argument is unnecessary. Set `labels='omit'` if there are embedded labels but you do not want them in the table.

`labels` can be used in any one of three formats.

### `labels` as a vector

`labels` can be set to be a vector of equal length to the number of variables in `data`, and in the same order. `NA` values can be used for padding if some variables do not have labels.

```{r, eval=FALSE}
library(vtable)

#Note that LifeCycleSavings has five variables
data(LifeCycleSavings)
#These variable labels are taken from help(LifeCycleSavings)
labs <- c('numeric aggregate personal savings',
    'numeric % of population under 15',
    'numeric % of population over 75',
    'numeric real per-capita disposable income',
    'numeric % growth rate of dpi')
vtable(LifeCycleSavings,labels=labs)

labs <- c('numeric aggregate personal savings',NA,NA,NA,NA)
vtable(LifeCycleSavings,labels=labs)
```
### `labels` as a two-column data set

`labels` can be set to a two-column data set (any type will do) where the first column has the variable names, and the second column has the labels. The column names don't matter.

This approach does __not__ require that every variable name in `data` has a matching label.

```{r, eval=FALSE}
library(vtable)

#Note that LifeCycleSavings has five variables
#with names 'sr', 'pop15', 'pop75', 'dpi', and 'ddpi'
data(LifeCycleSavings)
#These variable labels are taken from help(LifeCycleSavings)
labs <- data.frame(nonsensename1 = c('sr', 'pop15', 'pop75'),
nonsensename2 = c('numeric aggregate personal savings',
    'numeric % of population under 15',
    'numeric % of population over 75'))
vtable(LifeCycleSavings,labels=labs)

labs <- as.matrix(labs)
vtable(LifeCycleSavings,labels=labs)
```

### `labels` as a one-row data set

`labels` can be set to a one-row data set in which the column names are the variable names in `data` and the first row is the variable names. The `labels` argument can take any data type including data frame, data table, tibble, or matrix, as long as it has a valid set of variable names stored in the `colnames()` attribute.

This approach does __not__ require that every variable name in `data` has a matching label.

```{r, eval=FALSE}
library(vtable)

#Note that LifeCycleSavings has five variables
#with names 'sr', 'pop15', 'pop75', 'dpi', and 'ddpi'
data(LifeCycleSavings)
#These variable labels are taken from help(LifeCycleSavings)
labs <- data.frame(sr = 'numeric aggregate personal savings',
    pop15 = 'numeric % of population under 15',
    pop75 = 'numeric % of population over 75')
vtable(LifeCycleSavings,labels=labs)

labs <- as.matrix(labs)
vtable(LifeCycleSavings,labels=labs)
```

## `class`

The `class` flag will either report or not report the class of each variable in the resulting variable table. By default this is set to `TRUE`.

```{r, eval=FALSE}
library(vtable)

data(LifeCycleSavings)
vtable(LifeCycleSavings)
vtable(LifeCycleSavings,class=FALSE)
```

## `values`

The `values` flag will either report or not report the values that each variable takes. Numeric variables will report a range, logicals will report 'TRUE FALSE', and factor variables will report the first `factor.limit` (default 5) factors listed.

If the variable is numeric but has value labels applied by the `sjlabelled` package, `vtable()` will find them and report the numeric-label crosswalk.

```{r, eval=FALSE}
library(vtable)

data(LifeCycleSavings)
vtable(LifeCycleSavings,values=FALSE)
vtable(LifeCycleSavings)

#CO2 contains factor variables
data(CO2)
vtable(CO2)

#efc contains labeled values
#Note that the original value labels do not tell easily you what numerical
#value each label maps to, but vtable() does.
library(sjlabelled)
data(efc)
vtable(efc)
```

## `missing`

The `missing` flag, set to TRUE, will report whether or not the variable has any missing values. Defaults to FALSE.

```{r, eval=FALSE}
library(vtable)

data(LifeCycleSavings)
LifeCycleSavings$sr[1] <- NA
vtable(LifeCycleSavings,missing=TRUE)
```

## `index`

The `index` flag will either report or not report the index number of each variable. Defaults to FALSE.

```{r, eval=FALSE}
library(vtable)

data(LifeCycleSavings)
vtable(LifeCycleSavings,index=TRUE)
```

## `factor.limit`

If `values` is set to `TRUE`, then `factor.limit` limits the number of factors displayed on the variable table. `factor.limit` is by default 5, to cut down on clutter. The table will include the phrase "and more" to indicate that some factors have been cut off.

Setting `factor.limit=0` will include all factors.  If `values=FALSE`, `factor.limit` does nothing.

```{r, eval=FALSE}
library(vtable)

#CO2 contains factor variables
data(CO2)
vtable(CO2)
vtable(CO2,factor.limit=1)
vtable(CO2,factor.limit=0)
```

## `char.values`

If `values` is set to `TRUE`, then `char.values = TRUE` instructs `vtable` to list the values that character variables take, as though they were factors. If you only want some of the character variables to have their values listed, use a character vector to indicate which variables.

```{r, eval=FALSE}
library(vtable)

data(USJudgeRatings)
USJudgeRatings$Judge <- row.names(USJudgeRatings)
USJudgeRatings$SecondCharacter <- 'Less Interesting'
USJudgeRatings$ThirdCharacter <- 'Less Interesting Still!'

#Show values for all character variables
vtable(USJudgeRatings,char.values=TRUE)
#Or just for a subset
vtable(USJudgeRatings,char.values=c('Judge','SecondCharacter'))
```

## `slow.ok`

If the data contains labelled values, but the labels don't line up properly with the actual values in the data, that's a problem for `vtable` as it tries to match up data values with labels.

`vtable` has a fix for this but it's very slow, on the order of a few minutes for larger data sets. 

So in this case, by default (`slow.ok = FALSE`), the Values column will not attempt to link values to labels, and will just show the label for labelled values and raw data for unlabelled values. 

Set `slow.ok = TRUE` to instead get the correct result, slowly.

To avoid this being an issue at all, ensure that labels match values exactly before calling `vtable`, perhaps by running `sjlabelled::drop_labels()` and `sjlabelled::fill_labels()`.

## `data.title`

`data.title` will include a data title in the variable documentation file. If not set manually, this will default to the variable name for `data`.

```{r, eval=FALSE}
library(vtable)

data(LifeCycleSavings)
vtable(LifeCycleSavings)
vtable(LifeCycleSavings,data.title='Intercountry Life-Cycle Savings Data')
```

## `desc`

`desc` will include a description of the data set in the variable documentation file. This will by default include information on the number of observations and the number of columns. To remove this, set `desc='omit'`, or include any description and then include 'omit' as the last four characters. 

```{r, eval=FALSE}
library(vtable)

data(LifeCycleSavings)
vtable(LifeCycleSavings)
vtable(LifeCycleSavings,data.title='Intercountry Life-Cycle Savings Data',
    desc='Data on the savings ratio 1960–1970.')
vtable(LifeCycleSavings,data.title='Intercountry Life-Cycle Savings Data',
    desc='omit')
vtable(LifeCycleSavings,data.title='Intercountry Life-Cycle Savings Data',
    desc='Data on the savings ratio 1960–1970. omit')
```

## `col.width`

`vtable()` will select default column widths for the variable table depending on which measures `(name, class, label, values, summ)` are included. `col.width`, as a vector of percentage column widths on the 0-100 scale, will override these defaults.

```{r, eval=FALSE}
library(vtable)
library(sjlabelled)

data(efc)
vtable(efc)
#The variable names in this data set are pretty short, and the value labels are
#a little cramped, so let's move that over.
vtable(efc,col.width=c(10,10,40,40))
```

## `summ`

`summ` will calculate summary statistics for all non-missing values of each numeric and logical variable. `summ` is very flexible. It takes a character vector in which each element is of the form `function(x)`, where `function(x)` is any function that takes a vector and returns a single numeric value. For example, `summ=c('mean(x)','median(x)','mean(log(x))')` would calculate the mean, median, and mean of the log for each variable.

`summ` also takes two functions that are not R standards: `propNA(x)` and `countNA(x)`, which give the proportion and count of NA values in the variable, respectively. These two functions are always reported first, and are calculated for all variables, not just numeric and logical ones.

```{r, eval=FALSE}
library(vtable)
library(sjlabelled)
data(efc)

vtable(efc,summ=c('mean(x)','countNA(x)'))
```

-----

# The `labeltable()` function

`vtable()` syntax follows the following outline:

```{r, eval=FALSE}
labeltable(var,
    ...,
    out=NA,
    file=NA, 
    desc=NA)
```

`labeltable()` is a function that shows the values that correspond to `var`. This could be value label values, or it could be the values found in the data for the `...` variables.

```{r, eval=FALSE}
#Include a single labelled variable to show how the values of that variable correspond to its value labels.
library(sjlabelled)
data(efc)
labeltable(efc$e15relat)

#Include more than one variable to show, for each value of the first, what values of the others are present in the data.
data(mtcars)
labeltable(mtcars$cyl,mtcars$carb,mtcars$am)
```

## `out`

The `out` option determines what will be done with the resulting label table file. There are several options for `out`:

| Option     | Result |
|------------| -----------------------------------------|
| browser    | Loads variable documentation in web browser. |
| viewer     | Loads variable documentation in Viewer pane (RStudio only). |
| htmlreturn | Returns HTML code for variable documentation file. |
| return     | Returns variable documentation table in data frame format. |

## `file`

The `file` argument will write the variable documentation file to an HTML file and save it. Will automatically append 'html' filetype if the filename does not include a period.

```{r, eval=FALSE}
library(vtable)

data(LifeCycleSavings)
vtable(LifeCycleSavings,file='lifecycle_variabledocumentation')
```

## `desc`

`desc` will include a description of the data set in the variable documentation file, which may be useful for documentation purposes.

-----

# The `dftoHTML()` function

`dftoHTML()` syntax follows the following outline:

`dftoHTML(data,out=NA,file=NA,col.width=NA,row.names=FALSE)`

`dftoHTML()` largely exists to serve `vtable()`. What it does is takes a data set `data` and returns an HTML table with the contents of that data. 

Outside of its use in `vtable()`, `dftoHTML()` can also be used to keep a view of the data file open while working on the data, avoiding repeated calls to `head()` or similar, or switching back and forth between code tabs and data view tabs.

------

## `data`

`dftoHTML()` will accept any data set with a `colnames()` attribute.

```{r, eval=FALSE}
library(vtable)

data(LifeCycleSavings)
dftoHTML(LifeCycleSavings)
```

## out

The `out` option determines what will be done with the resulting variable documentation file. There are several options for `out`:

| Option     | Result |
|------------| -----------------------------------------|
| browser    | Loads HTML version of `data` in web browser. |
| viewer     | Loads HTML version of `data` in Viewer pane (RStudio only). |
| htmlreturn | Returns HTML code for `data`. |

By default, `vtable` will select 'viewer' if running in RStudio, and 'browser' otherwise.

```{r, eval=FALSE}
library(vtable)

data(LifeCycleSavings)
dftoHTML(LifeCycleSavings)
dftoHTML(LifeCycleSavings,out='browser')
dftoHTML(LifeCycleSavings,out='viewer')
htmlcode <- dftoHTML(LifeCycleSavings,out='htmlreturn')
```

## `file`

The `file` argument will write the HTML version of `data` to an HTML file and save it. Will automatically append 'html' filetype if the filename does not include a period.

```{r, eval=FALSE}
library(vtable)

data(LifeCycleSavings)
dftoHTML(LifeCycleSavings,file='lifecycledata_htmlversion.html')
```

## `col.width`

`dftoHTML()` will select, by default, equal column widths for all columns in `data`. `col.width`, as a vector of percentage column widths on the 0-100 scale, will override these defaults.

```{r, eval=FALSE}
library(vtable)

data(LifeCycleSavings)
dftoHTML(LifeCycleSavings)
#Let's make sr much bigger for some reason
dftoHTML(LifeCycleSavings,col.width=c(60,10,10,10,10))

```

## `row.names`

The `row.names` flag determines whether the row names of the data are included as the first column in the output table. 

```{r, eval=FALSE}
library(vtable)

data(LifeCycleSavings)
dftoHTML(LifeCycleSavings,row.names=TRUE)
```
