---
title: "vtable Bonus Functions"
author: "Nick Huntington-Klein"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
<!-- output: rmarkdown::html_vignette. pdf_document -->
vignette: >
  %\VignetteIndexEntry{vtable Bonus Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The `vtable` package serves the purpose of outputting automatic variable documentation that can be easily viewed while continuing to work with data.

`vtable` contains four main functions: `vtable()` (or `vt()`), `sumtable()` (or `st()`),  `labeltable()`, and `dftoHTML()`/`dftoLaTeX()`. 

This vignette focuses on some bonus helper functions that come with `vtable` that have been exported because they may be handy to you. This can come in handy for saving a little time, and can help you avoid having to create an unnamed function when you need to call a function.

-----

# Shortcut Helper Functions

`vtable` includes four shortcut functions. These are generally intended for use with the `summ` option in `vtable` and `sumtable` because nested functions don't look very nice in a `vtable`, or in a `sumtable` unless you explicitly set the `summ.names`. 

## `nuniq`

`nuniq(x)` returns `length(unique(x))`, the number of unique values in the vector.

## `countNA`, `propNA`, and `notNA`

These three functions are shortcuts for dealing with missing data. You have probably written out the nested versions of these many times!

| Function     | Short For |
|------------| -----------------------------------------|
| `countNA()`    | `sum(is.na())` |
| `propNA()`     | `mean(is.na())` |
| `notNA()` | `sum(!is.na())` |

## `is.round`

This function is a shortcut for `!any(!(x == round(x,digits)))`.

It takes two arguments: a vector `x` and a number of `digits` (0 by default). It checks whether you can round to `digits` digits without losing any information.

-----

# Other Helper Functions

## `pctile`

`pctile(x)` is short for `quantile(x,1:100/100)`. So in one sense this is another shortcut function. But this inherently lets you interact with percentiles a bit differently.

While `quantile()` has you specify which percentile you want in the function call, `pctile()` returns an object with all integer percentiles, and you can pull out which ones you want afterwards. `pctile(x)[50]` is the 50th percentile, etc.. This can be convenient in several applications, an obvious one being in `sumtable`.

```{r}
library(vtable)
#Some random normal data, and its percentiles
d <- rnorm(1000)
pc <- pctile(d)

#25th, 50th, 75th percentile
pc[c(25,50,75)]
```
```{r}
#Inverse normal CDF with 100 points of articulation
plot(pc)
```

# `independence.test`

`independence.test` is a helper function for `sumtable(group.test=TRUE)` that tests for independence between a categorical variable `x` and another variable `y` that may be categorical or numerical. 

Then, it outputs a *formatted string* as its output, with significance stars, for printing.

The function takes the format

```{r, eval = FALSE}
independence.test(x,y,
                  factor.test = NA,
                  numeric.test = NA,
                  star.cutoffs = c(.01,.05,.1),
                  star.markers = c('***','**','*'),
                  digits = 3,
                  fixed.digits = FALSE,
                  format = '{name}={stat}{stars}',
                  opts = list())
```

## `factor.test` and `numeric.test`

These are functions that actually perform the independence test. `numeric.test` is used when `y` is numeric, and `factor.test` is used in all other instances.

Specifically, these functions should take only `x` and `y` as arguments, and should return a list with three elements: the name of the test statistic, the test statistic itself, and the p-value of the test.

By default, these are the internal functions `vtable:::chisq.it` for `factor.test` and `vtable:::groupf.it` for `numeric.test`, so you can take a look at those (just put `vtable:::chisq.it` in the terminal and it will show you the function's code) if you'd like to make your own test functions.

## `star.cutoffs` and `star.markers`

These are numeric and character vectors, respectively, used for p-value cutoffs and to create significance markers. 

`star.cutoffs` indicates the cutoffs, and `star.markers` indicates the markers to be used with each cutoff, in the same order. So with `star.cutoffs = c(.01,.05,.1)` and `star.markers = c('***','**','*')`, each p-value below .01 will get marked with `'***'`, each from .01 to .05 will get `'**'`, and each from .05 to .1 will get `*`.

Defaults are set to "economics defaults" (.1, .05, .01). But these are of course easy to change.

```{r}
data(iris)
independence.test(iris$Species,
                  iris$Sepal.Length,
                  star.cutoffs = c(.05,.01,.001))
```

## `digits` and `fixed.digits`

`digits` indicates how many digits after the decimal place from the test statistics and p-values should be displayed. `fixed.digits` determines whether trailing zeros are maintained.

```{r}
independence.test(iris$Species,
                  iris$Sepal.Width,
                  digits=1)
```

```{r}
independence.test(iris$Species,
                  iris$Sepal.Width,
                  digits=4,
                  fixed.digits = TRUE)
```

## `format`

This is the printing format that the output will produce, incorporating the name of the test statistic `{name}`, the test statistic `{stat}`, the significance markers `{stars}`, and the p-value `{pval}`.

If your `independence.test` is heading out to another format besides being printed in the R console, you may want to add additional markup like `'{name}$={stat}^{stars}$'}` in LaTeX or `'{name}={stat}<sup>{stars}</sup>'` in HTML. If you do this, be sure to think carefully about escaping or not escaping characters as appropriate when you print!

```{r}
independence.test(iris$Species,
                  iris$Sepal.Width,
                  format = 'Pr(>{name}): {pval}{stars}')
```

## `opts`

You can create a named list where the names are the above options and the values are the settings for those options, and input it into `independence.test` using `opts=`. This is an easy way to set the same options for many `independence.test`s.
